An Approach for the Reduction of Boundary Effects in Time-Frequency Representations

### 2020-05-27 Notes/results.tex
update code

\paragraph{Conclusion}
Both previous experimental results combined with the theoretical asymptotic equation~\eqref{eq:cov.error.2} allow us to describe the influence of the noise variance and the size of the training dataset on the variance of the forecasting noise, which is empirically summarized as follows:
\begin{equation}
\gamma[N-1+\ell,N-1+\ell] \underset{K\to\infty}{\approx} \dfrac{\sigma^2}{K}g[\ell] .
\end{equation} 
where $g$ is a bounded positive function. The empirical result is coherent with the theoretical result provided by Theorem~\ref{th:error}.

@@

\begin{table}
\centering
\caption{Performance of the extension methods.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Algorithm} & \multicolumn{2}{c|}{MSE}  & \multirow{2}{41pt}{Computing time (sec.)} \\
   \cline{2-3} & Mean & Standard deviation & \\
   \hhline{|=#=|=|=|}
   {\sf SigExt} & $1.433\times 10^{-3}$ & $4.361\times 10^{-4}$ & $0.15$ \\
   \hline
   EDMD & $3.076\times 10^{-2}$ & $8.095\times 10^{-2}$ & $2.53$\\
   \hline
   GPR & $1.436\times 10^{-3}$ & $4.346\times 10^{-4}$ & $146.33$ \\
   \hline
   TBATS & ${\color{red}X}$ & ${\color{red}X}$ & ${\color{red}X}$ \\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

@@

In the aim of comparing the different techniques, we uses a criterion, proposed in~\cite{Daubechies16conceft}, that quantify the distance between a given time-frequency representation and the optimal one. It is built in analogy with the optimal transport distance, which enables quantifying the distance between two probability density functions. Let us generically denote a time frequency representation $\ccQ$. Then, for $t$ fixed, we consider the following pseudo-probability density function:
\begin{align}
p_\ccQ^t(\xi) &= \dfrac{|\ccQ(\xi,t)|^2}{\int_\RR |\ccQ(\nu,t)|^2\dd\nu}\ .
\label{eq:ppdf}
\end{align}
Then, at each instant $t$, we can then determine the optimal transport distance $d_{t}$ between the two pseudo densities. It is given by the $L^1$ norm of the difference between the associated distribution functions. In other words, we have:
\begin{equation*}
d_{t}(\ccQ,\ccF_0) = \int_\RR\left|\tilde P_{\ccQ}^t(\xi)-  P_{\ccF_0}^t(\xi)\right|\dd\xi\ ,
\end{equation*}
where $P_{\ccQ}^t(\xi)=\int_{-\infty}^\xi p_{\ccQ}^t(\nu)\dd\nu$ and $\tilde P_{\ccF_0}^t(\xi)=\int_{-\infty}^\xi\tilde p_{\ccF_0}^t(\nu)\dd\nu$.
Finally, the distance between the two time-frequency representations is obtained by averaging all the optimal transport distances with respect to time:
\begin{equation}
D(\ccQ,\ccF_0) = \frac1{|I|}\int_I d_{t}(\ccQ,\ccF_0)\dd t\ .
\end{equation}
The \textit{Optimal Transport Distance} (OTD) quantifies the proximity between the estimated and actual instantaneous frequencies while favouring the sparsity of the estimated time-frequency representation.

@@

\begin{table}
\centering
\caption{MSE of the respiratory signal extensions.}
\begin{tabular}{|c||c|c|}
  \hline
   \multirow{2}{*}{Algorithm} & \multicolumn{2}{c|}{MSE} \\
   \cline{2-3}
      & Mean & Median\\
   \hhline{|=#=|=|}
   {\sf SigExt} & $1.704 \times 10^{1}$ & $7.346 \times 10^{-1}$ \\
   \hline
   EDMD & $2.172\times 10^{-2}$ & $7.349\times 10^{-1}$ \\
   \hline
   GPR & $4.476\times 10^{-2}$ & $1.998\times 10^{-2}$ \\
   \hline
\end{tabular}
\label{tab:mse}
\end{table}

Then, we apply the boundary effects reduction method {\sf BoundEffRed} (Algorithm~\ref{alg:boundary}) on these extensions for diverse time-frequency representations (STFT, SST and RS). In Table~\ref{tab:otd.tho}, the results are compared in terms of OTD. They are also compared with the strategy consisting in a zero-padding extension of the signal. Even though the performance of the forecasting Algorithm~\ref{alg:extension} is somehow moderate, the boundary effects can be reduced dramatically on the time-frequency representations. Notice that the extension length $L$ has been set accordingly to the window length used by the time-frequency analysis tool. For instance, here the window length we use to evaluate the STFT is of 1500 samples. To prevent the STFT from being sensitive to the boundary effect, we set $L=750$. In this way, when evaluating the spectral content of the signal near its boundaries, the analysis is not limited by a lack of information all along the window support. From now on, all results are given for $L$ at equal to the half of the width of the window used in the time-frequency transform.

\begin{table}
\centering
\caption{Respiratory signal: OTD of the boundary effects reduction methods on diverse TFR.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Extension method} & \multicolumn{3}{c|}{Time-Frequency Representation} \\
   \cline{2-4}
      & STFT & SST & RS\\
   \hhline{|=#=|=|=|}
   Without extension & $2.16\times 10^{-2}$ & $5.26\times 10^{-3}$ & $3.07\times 10^{-2}$ \\
   \hline
   {\sf SigExt} & $1.72\times 10^{-2}$ & $4.00\times 10^{-3}$ & $2.43\times 10^{-2}$ \\
   \hline
   EDMD & $1.75\times 10^{-2}$ & $4.23\times 10^{-3}$ & $2.45\times 10^{-2}$ \\
   \hline
   GPR & $1.84\times 10^{-2}$ & $4.10\times 10^{-3}$ & $2.48\times 10^{-2}$ \\
   \hline
\end{tabular}
\label{tab:otd.tho}
\end{table}

@@

\begin{table}
\centering
\caption{Performance of the Boundary effect reduction on TFR.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Extension method} & \multicolumn{3}{c|}{Time-Frequency Representation} \\
   \cline{2-4}
      & STFT & SST & RS\\
   \hhline{|=#=|=|=|}
   Without extension & $2.52\times 10^{-2}$ & $9.41\times 10^{-2}$ & $1.03\times 10^{-1}$ \\
   \hline
   {\sf SigExt} & $1.22\times 10^{-3}$ & $7.31\times 10^{-2}$ & $1.11\times 10^{-1}$ \\
   \hline
   EDMD & $9.83\times 10^{-4}$ & $5.59\times 10^{-2}$ & $9.80\times 10^{-2}$ \\
   \hline
   GPR & $1.14\times 10^{-3}$ & $7.02\times 10^{-2}$ & $1.07\times 10^{-1}$ \\
   \hline
\end{tabular}
\label{tab:otd.ppg}
\end{table}

@@

>>>>

\paragraph{Summary}
Both previous experimental results combined with the theoretical asymptotic equation~\eqref{eq:cov.error.2} allow us to describe the influence of the noise variance and the size of the training dataset on the variance of the forecasting noise, which is empirically summarized as follows:
\begin{equation}
\gamma[N-1+\ell,N-1+\ell] \underset{K\to\infty}{\approx} \dfrac{\sigma^2}{K}g[\ell] .
\end{equation} 
where $g$ is a bounded positive function. The empirical result is coherent with the theoretical result provided by Theorem~\ref{th:error}.

@@

\begin{table}
\centering
\caption{AHM signal. Performance of the extension methods.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Algorithm} & \multicolumn{2}{c|}{MSE}  & \multirow{2}{41pt}{Computing time (sec.)} \\
   \cline{2-3} & Mean & Standard deviation & \\
   \hhline{|=#=|=|=|}
   {\sf SigExt} & $1.433\times 10^{-3}$ & $4.361\times 10^{-4}$ & $0.15$ \\
   \hline
   EDMD & $3.076\times 10^{-2}$ & $8.095\times 10^{-2}$ & $2.53$\\
   \hline
   GPR & $1.436\times 10^{-3}$ & $4.346\times 10^{-4}$ & $146.33$ \\
   \hline
   TBATS & ${\color{red}X}$ & ${\color{red}X}$ & ${\color{red}X}$ \\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

@@

In the aim of comparing the different techniques, we use a criterion, proposed in~\cite{Daubechies16conceft}, that quantify the distance between a given time-frequency representation and the optimal one. It is built in analogy with the optimal transport distance, which enables quantifying the distance between two probability density functions. Let us generically denote a time frequency representation $\ccQ$. Then, for $t$ fixed, we consider the following pseudo-probability density function:
\begin{align}
p_\ccQ^t(\xi) &= \dfrac{|\ccQ(\xi,t)|^2}{\int_\RR |\ccQ(\nu,t)|^2\dd\nu}\ .
\label{eq:ppdf}
\end{align}
Then, at each instant $t$, we can then determine the optimal transport distance $d_{t}$ between the two pseudo densities. It is given by the $L^1$ norm of the difference between the associated distribution functions. In other words, we have:
\begin{equation*}
d_{t}(\ccQ,\ccF_0) = \int_\RR\left|\tilde P_{\ccQ}^t(\xi)-  P_{\ccF_0}^t(\xi)\right|\dd\xi\ ,
\end{equation*}
where $P_{\ccQ}^t(\xi)=\int_{-\infty}^\xi p_{\ccQ}^t(\nu)\dd\nu$ and $\tilde P_{\ccF_0}^t(\xi)=\int_{-\infty}^\xi\tilde p_{\ccF_0}^t(\nu)\dd\nu$.
Finally, the distance between the two time-frequency representations is obtained by averaging all the optimal transport distances with respect to time:
\begin{equation}
D(\ccQ,\ccF_0) = \frac1{|I|}\int_I d_{t}(\ccQ,\ccF_0)\dd t\ .
\end{equation}
The \textit{Optimal Transport Distance} (OTD) quantifies the proximity between the estimated and actual instantaneous frequencies while favouring the sparsity of the estimated time-frequency representation.

@@

\begin{table}
\centering
\caption{Respiratory signal. MSE of the respiratory signal extensions.}
\begin{tabular}{|c||c|c|}
  \hline
   \multirow{2}{*}{Algorithm} & \multicolumn{2}{c|}{MSE} \\
   \cline{2-3}
      & Mean & Median\\
   \hhline{|=#=|=|}
   {\sf SigExt} & $1.704 \times 10^{1}$ & $7.346 \times 10^{-1}$ \\
   \hline
   EDMD & $2.172\times 10^{-2}$ & $7.349\times 10^{-1}$ \\
   \hline
   GPR & $4.476\times 10^{-2}$ & $1.998\times 10^{-2}$ \\
   \hline
\end{tabular}
\label{tab:mse}
\end{table}

Then, we apply the boundary effects reduction method {\sf BoundEffRed} (Algorithm~\ref{alg:boundary}) on these extensions for diverse time-frequency representations: STFT, SST, RS, as well as ConceFT, an SST-based representation introduced in~\cite{Daubechies16conceft}. In Table~\ref{tab:otd.tho}, the results are compared in terms of OTD. They are also compared with the strategy consisting in a zero-padding extension of the signal. Even though the performance of the forecasting Algorithm~\ref{alg:extension} is somehow moderate, the boundary effects can be reduced dramatically on the time-frequency representations. Notice that the extension length $L$ has been set accordingly to the window length used by the time-frequency analysis tool. For instance, here the window length we use to evaluate the STFT is of 1500 samples. To prevent the STFT from being sensitive to the boundary effect, we set $L=750$. In this way, when evaluating the spectral content of the signal near its boundaries, the analysis is not limited by a lack of information all along the window support. From now on, all results are given for $L$ at equal to the half of the width of the window used in the time-frequency transform.

\begin{table}
\centering
\caption{Respiratory signal. Averaged OTD of the boundary effects reduction methods on diverse representations.}
\begin{tabular}{|>{\centering}m{35pt}||c|c|c|c|}
  \hline
   \multirow{2}{35pt}{\centering Extension method} & \multicolumn{4}{c|}{Time-Frequency Representation} \\
   \cline{2-5}
      & STFT & SST & RS & ConceFT \\
   \hhline{|=#=|=|=|=|}
   Without extension & $2.16\times 10^{-2}$ & $5.26\times 10^{-3}$ & $3.07\times 10^{-2}$ & $1.41\times 10^{-2}$ \\
   \hline
   {\sf SigExt} & $1.72\times 10^{-2}$ & $4.00\times 10^{-3}$ & $2.43\times 10^{-2}$ & $1.12\times 10^{-2}$ \\
   \hline
   EDMD & $1.75\times 10^{-2}$ & $4.23\times 10^{-3}$ & $2.45\times 10^{-2}$ & {\color{red}X} \\
   \hline
   GPR & $1.84\times 10^{-2}$ & $4.10\times 10^{-3}$ & $2.48\times 10^{-2}$ & {\color{red}X} \\
   \hline
\end{tabular}
\label{tab:otd.tho}
\end{table}

@@

\begin{table}
\centering
\caption{PPG signal. Averaged OTD of the boundary effects reduction methods on diverse representations.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Extension method} & \multicolumn{3}{c|}{Time-Frequency Representation} \\
   \cline{2-4}
      & STFT & SST & RS\\
   \hhline{|=#=|=|=|}
   Without extension & $2.52\times 10^{-2}$ & $9.41\times 10^{-2}$ & $1.03\times 10^{-1}$ \\
   \hline
   {\sf SigExt} & $1.22\times 10^{-3}$ & $7.31\times 10^{-2}$ & $1.11\times 10^{-1}$ \\
   \hline
   EDMD & $9.83\times 10^{-4}$ & $5.59\times 10^{-2}$ & $9.80\times 10^{-2}$ \\
   \hline
   GPR & $1.14\times 10^{-3}$ & $7.02\times 10^{-2}$ & $1.07\times 10^{-1}$ \\
   \hline
\end{tabular}
\label{tab:otd.ppg}
\end{table}

@@

To evaluate the influence of the noise level on the performance of {\sf BoundEffRed}, we artificially add a Gaussian noise to the measured PPG signal, whose variance is denoted $\sigma_{\mathrm{add}}^2$. It is thus an additional noise to the measurement noise actually contained in the signal. Figure~\ref{fig:otd.noise} shows the averaged OTD of {\sf BoundEffRed} for different values of $\sigma_{\mathrm{add}}^2$. At first glance, we notice that STFT is slightly more sensitive to noise than SST. On one hand, the robustness to noise of the SST is the direct consequence of the approximation result of~\cite{Daubechies16conceft}, discussed in section~\ref{sse:perf.BoundEffRed}. On the other hand, the sentivity to noise of the STFT may be interpreted as the inability to forecast the values of the time-frequency coefficients where only noise is active, while sharp time-frequency representations, such as SST, ideally vanish at these points.
\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{OTDBoundEffRed.eps}
\caption{PPG signal. Averaged OTD of {\sf BoundEffRed} in function of the noise variance.}
\label{fig:otd.noise}
\end{figure}
